---
pre: <b>10/5. </b>
title: "Dynamic Plotting"
weight: 13
summary: "How to elevate your plots."
format:
    hugo:
      toc: true
      output-file: "_index.en.md"
      reference-links: true
      code-link: true
      
---

```{r, include=FALSE}
answers = FALSE

library(ggplot2)
```

## Overview




## The Data

Today we will be using data from the Environmental Protection Agency (EPA) [Fuel Economy data](https://fueleconomy.gov/). It comes with the `ggplot2` package, and we can load it into the environment using the following (only *after* you have loaded `ggplot2`):

```{r}
library(ggplot2)

mpg = data.frame(mpg)
```

Here is a rundown of what the variables are:

manufacturer
:   manufacturer name

model
:   model name

displ
:   engine displacement, in liters

year
:   year of manufacture

cyl
:   number of cylinders

trans
:   type of transmission

drv
:   the type of drive train, where f = front-wheel drive, r = rear wheel drive, 4 = 4wd

cty
:   city miles per gallon

hwy
:   highway miles per gallon

fl
:   fuel type

class
:   "type" of car

## Problem Sets

### 0. Make a Plot

Before we start making a fancy plot, we need a basic plot. I will provide the code for one here so everyone is starting from the same place.

```{r}
ggplot(mpg) +
  aes(x = displ, y = hwy) +
  geom_point(size = 1.5) +
  labs(
    x = "Engine Displacement",
    y = "Highway MPG",
    title = "Highway MPG by Engine Size",
    caption = "Source: https://fueleconomy.gov/"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold"))
```

::: {.question}
We will be modifying this plot for the remainder of the worksheet. Make sure you understand all the parts! 
:::

### 1. Adding Labels to a ggplot

In plot terminology, labels are text we add to a plot to explain some element; we add labels to the axes for example. We can also add labels to the data itself if it will help us tell our story. It's easy to go too far though, so this is often used sparingly.

There are several methods to add labels, but I'm going to show you the most general one. We start by adding a new column to our dataframe, which we'll call `labels`.

::: {.question}
Add a new `labels` column to our `mpg` dataframe. For now, just copy the `model` column
:::

```{r, include=answers, results='asis', echo=FALSE}
cat("
::: {.answer}
mpg$labels = mpg$model
:::
")
```

Let's see how we can incorporate these labels. 

::: {.question}
Modify the plot above by adding a new argument to the `aes()` function called `label`, and provide it our new `labels` column. Then, add a new gemom, `geom_text()`, to the ggplot code. In the text geom, include the argument `vjust = -1` to move the labels a bit so they aren't right on top of our points (which makes them hard to read).
:::

```{r, include=answers, results='asis', echo=FALSE}
cat('
::: {.answer}
ggplot(mpg) +
  aes(x = displ, y = hwy, label = labels) +
  geom_point(size = 1.5) +
  labs(
    x = "Engine Displacement",
    y = "Highway MPG",
    title = "Highway MPG by Engine Size",
    caption = "Source: https://fueleconomy.gov/"
  ) +
  geom_text(vjust = -1) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold"))
:::
')
```

Well, it worked! Not the nicest to look at though. We need to be selective about where we add our labels. Let's cut back a bit.

::: {.question}
Fill the `lables` column with `NA`s. Then, use sub-setting to add a label only for rows where the `class` is "midsize."
:::

```{r, include=answers, results='asis', echo=FALSE}
cat('
::: {.answer}
mpg$labels = NA

mpg$labels = ifelse(mpg$class == "midsize", "Midsize", NA)
:::
')
```

### 2. Isolating Elements

We're getting closer. Adding text to a plot can he helpful, but only for a small handful of points. If we want to highlight a whole group of points, some other method is typically more helpful. Let's try using color to highlight our midsize cars. We can do this by adding another layer on top of our current points, but only for those we want to highlight.

::: {.question}
Remove the `label` argument and text geom we added in the last section. Next, add a new `geom_point()` below the one we already have.
:::

```{r, include=answers, results='asis', echo=FALSE}
cat('
::: {.answer}

:::
')
```












### 3. Make it Interactive




### 4. Make it Accessible










